---
title: "R Exercise"
author: "Héctor Corrada Bravo"
date: '`r Sys.Date()`'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this exercise you will create your first submission using R Markdown. We will see more information about R Markdown in class, but you can also take a look below for some preliminary information.

Fill in your answers to the five questions below in the format requested (see example). Knit the file into a PDF and submit on ELMS.

## Loading data

You can use the following snippet, which we also used in the lecture notes used in class:

```{r download, echo=FALSE}
if (!file.exists("BPD_Arrests.csv")) {
  download.file("http://www.hcbravo.org/IntroDataSci/misc/BPD_Arrests.csv", destfile="BPD_Arrests.csv")
}

arrest_tab <- read.csv("BPD_Arrests.csv", stringsAsFactors=FALSE)

colnames(arrest_tab)[c(3,4)] <- c("race","sex")

arrest_tab$race <- factor(arrest_tab$race)
arrest_tab$sex <- factor(arrest_tab$sex)
arrest_tab$incidentOffense <- factor(arrest_tab$incidentOffense)
```
## Example

Use this format for each of the questions below: 
- question: The question, which we provide
- code: The code you used to answer it in an R code chunk
- statement: A succint prose statement (use complete sentences please) of the answer to the question.

**Question 0**: What is the minimum arrest age in the dataset?

```{r q0}
min(arrest_tab$age)
```

The minimum arrest age is 0.

## The questions

**Question 1**: What is the age of the oldest person arrested in the dataset?

```{r q1}
res <- max(arrest_tab$age)
res
```

The age of the oldest person is `r res`.

**Question 2**: What is the neighborhood with the least number of arrests?

```{r q2}
tmp <- table(arrest_tab$neighborhood)
tmp <- sort(tmp)
res <- tmp[1]
res
```

The neighborhood with the least number of arrests is `r res`.

**Question 3**: What is the median arrest age?

```{r q3}
res <- median(arrest_tab$age)
res
```

The median arrest age is `r res`

**Question 4**: What is the median arrest age in the SOUTHERN district?

```{r q4}
index_vector <- arrest_tab$district == "SOUTHERN"
arrest_subset <- arrest_tab[index_vector,]
res <- median(arrest_subset$age)
res
```

The median arrest age in the SOUTHERN district is `r res`.

**Question 5**: What is the mean age of females arrested in the “Federal Hill” neighborhood?

```{r q5}
index_vector <- arrest_tab$neighborhood == "Federal Hill" & 
                arrest_tab$sex == "F"
arrest_subset <- arrest_tab[index_vector,]
res <- mean(arrest_subset$age)
res
```

The mean age of females arrested in the Federal Hill neighborhood is `r res`.

