---
title: 'HW1: Datatypes and Wrangling'
author: "<Your Name Here>"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_notebook: default
editor_options:
  chunk_output_type: inline
---

```{r knitr_setup, echo=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

## Data types

_1) Provide a URL to the dataset._


_2) Explain why you chose this dataset._ 


_3) What are the entities in this dataset? How many are there?_


_4) How many attributes are there in this dataset?

_5) What is the datatype of each attribute (categorical -ordered or unordered-, numeric -discrete or continuous-, datetime, geolocation, other)? Write a short sentence stating how you determined the type of each attribute. Do this for at least 5 attributes, if your dataset contains more than 10 attributes, choose 10 of them to describe._

| Num | Name | Type | Description |
|-----|------|------|-------------|
| 1 | <name> | <type> | <descr> |

_6) Write R code that loads the dataset using function `read_csv`. Were you able to load the data successfully? If no, why not?_

```{r load_data, message=FALSE}
library(tidyverse)
# loading code goes here
```

## Wrangling

1) My pipeline computes... 

```{r pipeline}
# pipeline goes here
```

## Plotting

1) This plot shows... 

```{r plot}
# plot goes here
```