---
title: 'HW1: Datatypes and Wrangling'
author: "Hector Corrada Bravo"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_notebook: default
editor_options:
  chunk_output_type: inline
---

```{r knitr_setup, echo=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

## Data types

_1) Provide a URL to the dataset._

I downloaded my dataset from http://www.hcbravo.org/IntroDataSci/misc/BPD_Arrests.csv

_2) Explain why you chose this dataset._ 

I am interested in studying how rates of arrests in different parts of Baltimore are related to demographic statistics.

_3) What are the entities in this dataset? How many are there?_

Entities are specific arrests. There are 104528.

_4) How many attributes are there in this dataset?_

There are 15 attributes.

_5) What is the datatype of each attribute (categorical -ordered or unordered-, numeric -discrete or continuous-, datetime, geolocation, other)? Write a short sentence stating how you determined the type of each attribute. Do this for at least 5 attributes, if your dataset contains more than 10 attributes, choose 10 of them to describe._

| Num | Name | Type | Description |
|-----|------|------|-------------|
| 1 | `arrest` | categorical | Identifier of each arrest, takes values from finite set |
| 2 | `age` | numeric continuous | Ages are numeric values measured in time units |
| 3 | `race` | categorical unordered | Can take value from finite set of possible races |
| 4 | `sex` | categorical unordered | Can take value from finite set of possible sexes |
| 5 | `arrestDate` | datetime | Specifies date of arrest |
| 6 | `arrestTime` | datetime | Specifies time of arrest |
| 7 | `arrestLocation` | other - address | Street address of arrest |
| 8 | `incidentOffense` | categorical unordered | Can take value from finite set of possible offenses |
| 9 | `incidentLocation` | other - address | Stree address if incident |
| 10 | `charge` | categorical unordered | Can take value from finite set of possible charges |

_6) Write R code that loads the dataset using function `read_csv`. Were you able to load the data successfully? If no, why not?_

```{r load_data, message=FALSE}
library(tidyverse)

url <- "http://www.hcbravo.org/IntroDataSci/misc/BPD_Arrests.csv"
arrest_tab <- read_csv(url)
arrest_tab %>% slice(1:10)
```

## Wrangling

1) My pipeline computes average arrest age (ignoring ages <= 0), for each district and writes them in increasing order. It would be useful to see which districts tend to arrest younger individuals.

```{r pipeline}
mean_ages <- arrest_tab %>%
  filter(age > 0) %>%
  select(district, age) %>%
  group_by(district) %>%
  summarize(mean_age=mean(age)) %>%
  arrange(mean_age)
mean_ages
```

## Plotting

1) This barplot shows the average arrest age per district (ignoring ages <= 0)

```{r plot}
mean_ages %>%
  ggplot(aes(x=district, y=mean_age)) +
    geom_bar(stat="identity") +
    coord_flip()
```