# DB Parting Shots

## Database Query Optimization

Earlier we made the distinction that SQL is a _declarative_ language rather than a _procedural_ language. A reason why data base systems rely on a declarative language is that it allows the system to decide how to _evaluate_ a query _most efficiently_. Let's think about this briefly.

Consider a database where we have two tables `Batting` and `Master` and we want to evaluate this query: that _what is the maximum batting "average" for a player from the state of California_?

```{r, echo=FALSE}
db <- DBI::dbConnect(RSQLite::SQLite(), "data/lahman2016.sqlite")
```

```{sql, connection=db}
select max(1.0 * b.H / b.AB) as best_ba
from Batting as b join Master as m on b.playerId = m.playerId
where b.AB >= 100 and m.birthState = "CA"
```


Now, let's do the same computation using `dplyr` operations:

```{r}
library(Lahman)
library(tidyverse)
```

Here is one version that joins the two tables before filtering the rows that are included in the result.

```{r}
Batting %>%
  inner_join(Master, by="playerID") %>%
  filter(AB >= 100, birthState == "CA") %>%
  mutate(AB=1.0 * H / AB) %>%
  summarize(max(AB))
```

Here is a second version that filters the rows of the tables _before_ joining the two tables.

```{r}
Batting %>%
  filter(AB >= 100) %>%
  inner_join(
    Master %>% filter(birthState == "CA")
  ) %>%
  mutate(AB = 1.0 * H / AB) %>%
  summarize(max(AB))
```

They both give the same result of course, but which one should be more efficient?

In SQL we only write the one query describing our desired result, with the _procedural_ versions with `dplyr` we need to think which of the two versions is more efficient.

Database systems use _query optimization_ to decide how to evaluate query efficiently. The goal of query optimization is to decide the most efficient query _plan_ to use to evaluate a query out of the many possible candidate plans it could use. It needs to solve two problems: search the space of possible plans, approximate the _cost_ of evaluating a specific plan. Let's ignore the first, and discuss briefly the second.

We should think of the two procedural versions above as two candidate plans that the DB system _could_ use to evaluate the query. Query optimzation _approximates_ what it would cost to evaluate each of the two plans and decides to use the most efficient plan. 


So, how does it approximate cost? A few ingredients are used:

- Access cost: how much will it cost to access rows that satisfy a given predicate (`where` clause)? Consider the `Master` table. In our query we only need to find rows for players born in California. Suppose we have an _index_ based on attribute `birthState`, e.g. a hash table that allows us to find rows for players from a specific state very efficiently. In that case, accessing these rows using the index is much more efficient than scanning the entire table. This is why creating indexes for tables becomes important.

- Operation cost: how much will it cost to perform a join? There is a difference between comparing every pair of rows in order to compute a join, versus using indexes to find a small number of rows that satisfy the join condition efficiently? For example, if the `Batting` table has an index on `playerId` it will be cheaper to join with a filtered `Master` table, i.e., only considering rows for players born in California.

- Result size estimation: how many rows will we get after we perform a join? We can use information on key constraints to estimate this type of result. Additionally, these estimates also depend on the number of rows that satisfy certain predicates (e.g., number of players born in California) so systems often use histograms to make these estimates.

As database system users we may create indices or key constraints that guide the query optimizer to choose more efficient queries.

```{r, echo=FALSE}
DBI::dbDisconnect(db)
```

## JSON Data Model

The Entity-Relational data model we have described so far is mostly defined for _structured data_: where a specific and consistent schema is assumed.

Data models like XML and JSON are instead intended for *semi-structured* data.

#### XML: eXtensible Markup Language 

Data models like XML rely on flexible, self-describing schemas: 

```xml
<?xml version="1.0" encoding="UTF-8"?>
<!-- Edited by XMLSpy -->
<CATALOG>
  <CD>
    <TITLE>Empire Burlesque</TITLE>
    <ARTIST>Bob Dylan</ARTIST>
    <COUNTRY>USA</COUNTRY>
    <COMPANY>Columbia</COMPANY>
    <PRICE>10.90</PRICE>
    <YEAR>1985</YEAR>
  </CD>
  <CD>
    <TITLE>Hide your heart</TITLE>
    <ARTIST>Bonnie Tyler</ARTIST>
    <COUNTRY>UK</COUNTRY>
    <COMPANY>CBS Records</COMPANY>
    <PRICE>9.90</PRICE>
    <YEAR>1988</YEAR>
  </CD>
  ...
```

#### JSON: Javascript Object Notation
        
Very similar to XML and seems to be replacing it for many purposes

```json
{
  "firstName": "John",
  "lastName": "Smith",
  "isAlive": true,
  "age": 25,
  "height_cm": 167.6,
  "address": {
    "streetAddress": "21 2nd Street",
    "city": "New York",
    "state": "NY",
    "postalCode": "10021-3100"
  },
  "phoneNumbers": [
    {
      "type": "home",
      "number": "212 555-1234"
    },
    {
      "type": "office",
      "number": "646 555-4567"
    }
  ],
  "children": [],
  "spouse": null
}
```

This is the format most contemporary data REST APIs use to transfer data. For instance, here is part of a JSON record from a Twitter stream:

```json
{
  "created_at":"Sun May 05 14:01:34+00002013",
  "id":331046012875583488,
  "id_str":"331046012875583488",
  "text":"\u0425\u043e\u0447\u0443, \u0447\u0442\u043e\u0431 \u0442\u044b \u0441\u0434\u0435\u043b\u0430\u043b \u0432\u0441\u0451 \u043d\u0435\u043e\u0431\u0445\u043e\u0434\u0438\u043c\u043e\u0435.\n \\,,\\ *_* \/,,\/",
  "source":"\u003ca href=\"http:\/\/twitterfeed.com\"rel=\"nofollow\"\u003etwitterfeed\u003c\/a\u003e",
  "in_reply_to_user_id_str":null,
  "user":{
    "id":548422428,
    "id_str":"548422428",
    "name":"\u0410\u0439\u0433\u0435\u0440\u0438\u043c \u041f\u043e\u0433\u043e\u0434\u0438\u043d\u0430",
    "screen_name":"paddybyrny",
    "location":"\u0420\u043e\u0441\u0441\u0438\u044f;\u0412\u043b\u0430\u0434\u0438\u0432\u043e\u0441\u0442\u043e\u043a",
    "followers_count":4188,
    "friends_count":4281,
    "lang":"en",
    "profile_background_image_url":"http:\/\/a0.twimg.com\/images\/themes\/theme1\/bg.png",
  },
  "geo":null,
  "coordinates":null,
  "entities":{
    "hashtags":[],"symbols":[],"urls":[],"user_mentions":[]
  },"favorited":false,"retweeted":false,"filter_level":"medium","lang":"ru"}
```
