---
title: "EDA Activity"
author: "CMSC320"
date: "October 3, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Let's practice some EDA work. We're using the `Wage` dataset provided by the `ISLR` package. 

```{r}
library(tibble)
library(dplyr)
library(ggplot2)

library(ISLR)
data(Wage)

wage <- as_tibble(Wage)
wage
```

Let's warmup with one question:

**Q0**: How are wages distributed overall across years?

```{r, fig.width=3, fig.height=3, fig.align="center"}
wage %>%
  ggplot(aes(x=factor(year), y=wage, color=education)) +
    geom_boxplot()
```

Now, on your own:

**Q1**: How are wages distributed across years as a function of education? (Write the code to make this plot)

```{r, echo=FALSE, fig.width=6, fig.height=3, fig.align="center"}
wage %>%
  ggplot(aes(x=factor(year), y=wage, color=education)) +
    geom_boxplot()
```

**Q2**: How is the central tendency (e.g., median) of wage changing across years?

```{r}
# transform data
wage %>%
  group_by(year, education) %>%
  summarize(med_wage=median(wage)) %>%

# make the plot
  ggplot(aes(x=year,y=med_wage, color=education)) +
  geom_line()

```

**Q3**: How is median wage changing across years as a function of education?

```{r}
# transform data
wage %>%
  group_by(year) %>%
  summarize(med_wage=median(wage)) %>%

# make the plot
  ggplot(aes(x=,y=)) +
  geom_()

```

**Q4**: Is the wage gap between those with advanced degrees and those with less than a HS education changing over time?

_Part 1_: How are you going to define the wage gap? 

_Part 2_: Make a data frame with columns `year` and `wage_gap`.

```{r}
wage %>%
  group_by(year, education) %>%
  summarize(med_wage = median(wage)) %>%
  filter(str_detect(education, "^1") | str_detect(education, "^5")) %>%
  ungroup() %>%
  spread(education, med_wage) %>%
  select(year, hs=matches("^1"), ad=matches("^5")) %>%
  mutate(gap = (1 - (hs / ad)) * 100) %>%
  ggplot(aes(x=year, y=gap)) +
    geom_line()
```
_Part 3_: Plot wage gap as a function of year.