---
title: "Wrangling Exercise"
author: "CMSC320"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

This is an exercise on data wrangling using the flights dataset:

```{r, eval=TRUE, message=FALSE}
library(tidyverse)
library(nycflights13)
data(flights)
data(airlines)
```

```{r, echo=FALSE, fig.width=3, fig.height=3}
library(datamodelr)
dm_f <- dm_from_data_frames(flights, airlines)

dm_f <- dm_add_references(dm_f,
                          flights$carrier == airlines$carrier)

graph <- dm_create_graph(dm_f, rankdir="LR", col_attr = c("column", "type"))
dm_render_graph(graph)
```

1. Filter flights to include only flights with an arrival delay greater than 2 hours (delays are recorded in minutes)  
2. Select columns in the flights table that contain delay times (note that the function `matches` can be used to select columns with names that matches a regular expression. See `?dplyr::select`)  

3. Find the minimum arrival delay (this can be a negative number) in the flights table. Use `summarize`.  

4. Find minimum arrival delay (again, can be a negative number) in the flights table for each destination airport. Use `group_by' and 'summarize`.  

5. List the name of all airlines and the number of flights for each airline in flights table. The `airlines` table has airline names.



