---
title: "A short note about pipes"
author: "CMSC320"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Getting started

Let's load the same data we used in the previous lab (look there for instructions on how to download).

```{r}
arrest_tab <- read.csv("BPD_Arrests.csv", stringsAsFactors=FALSE)

# remember there's a naming issue in this dataset we need to fix
colnames(arrest_tab)[3:4] <- c("race", "sex")

arrest_tab$race <- factor(arrest_tab$race)
arrest_tab$sex <- factor(arrest_tab$sex)
arrest_tab$incidentOffense <- factor(arrest_tab$incidentOffense)
```

## Making the process explicit with pipes

We've discussed the idea of thinking about data analysis work in terms of "pipelines", where
we start from data of a certain shape (e.g., a `data.frame`) and apply transformations (functions) to obtain data that contains the computation we want. Consider the following example seen in class:

_What is the mean age of males arrested in the SOUTHERN district?_

We can frame the answer to this question as a series of data transformations to get the answer we are looking for:

```{r}
# filter data to observations we need
index_vector <- arrest_tab$sex == "M" & arrest_tab$district == "SOUTHERN"
tmp <- arrest_tab[index_vector,]

# select the attribute/column we need
tmp <-tmp[,"age"]

# compute statistic required
mean(tmp)
```

Let's rewrite this using functions to illustrate the point

```{r}
filter_data <- function(data) {
  index_vector <- data$sex == "M" & data$district == "SOUTHERN"
  data[index_vector,]
}

select_column <- function(data, column) {
  data[, column]
}

tmp <- filter_data(arrest_tab)
tmp <- select_column(tmp, "age")
mean(tmp)
```

So, this pattern of _data --transform--> data_ becomes clearer when written that way.

The `magrittr` package introduces _syntactic sugar_ to make this explicit. We can write the above snippet using the "pipe" operator `%>%`:

```{r}
library(magrittr)

arrest_tab %>%
  filter_data() %>%
  select_column("age") %>%
  mean()
```

The `%>%` binary operator takes the value to its **left** and inserts it as the first argument of the function call to its **right**. So the expression `LHS %>% f(another_argument)` is **equivalent** to the expression `f(LHS, another_argument)`. We will see this pattern extensively in class because it explicitly presents the way we want to organize many of our data analysis tasks.

