---
title: "Introduction to R"
author: "CMSC320"
date: "February 2, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

R is a computational environment for data analysis. It is designed around a _functional_ language, as opposed to _procedural_ languages like Java or C, that has desirable properties for the type of operations and workflows that are frequently performed in the course of analyzing datasets. In this exercise we will start learning some of those desirable properties while performing an analysis of a real dataset.

## Some history

R is an offspring of S, a language created in AT&T Labs by John Chambers (now at Stanford) and others in 1976 with the goal of creating an environment for statistical computing and data analysis. The standard for the language in current use was settled in 1998. That same year, "S" won the ACM Software System award, awarded to software systems "that have a lasting influence, reflected in contributions to concepts, in commercial acceptance, or both".

In 1991, Robert Gentleman and Ross Ihaka created R to provide an open source implementation of the S language and environment. They also redesigned the language to enforce lexical scoping rules. It has been maintained by the R core group since 1997, and in 2015 an R consortium, including Microsoft, Google, and others, was created.

## Availability and eco-system

R is freely available at [http://www.r-project.org/](http://www.r-project.org/). A number of good IDEs are available for its use:

- [Rstudio](http://rstudio.org)  
- [Revolution R](http://www.revolutionanalytics.com/get-revolution-r)  

You can also find bindings for commonly used editors and IDEs, like [emacs](http://ess.r-project.org/) and [eclipse](http://www.walware.de/goto/statet).

Resources for learning and reading about R are listed in our [course page](http://www.hcbravo.org/IntroDataScience/resources.html). Of note are the [swirl project](http://swirlstats.com/) and Data Camp's [introduction to R] course.

One of the biggest strengths of the R ecosystem is the variety and quality of packages for data analysis available. R uses a package system (like Python and Ruby for instance). Packages are divided into two classes: **base** which are packages installed when R is installed, includes packages for basic statistics, computing with probability distributions, plotting and graphics, matrix manipulations and other), all other packages are available in [CRAN](http://cran.r-project.org). We will be using a fair number of these packages through the course of the semester.

## How to think about R

As we wrote previously, R is an environment for data analysis, centered around a _functional_ language designed for common operations in data analysis. When we say that R is a _functional_ language, at its most basic definition, means that we treat _functions_ as values. For instance, we can assign _functions_ to variables, or pass _functions_ as arguments to other _functions_. We'll see examples of that later on. If you are familiar with Ruby or Python, while they are not functional languages, they do have some idioms of functional languages that may be familiar: [http://www.sitepoint.com/functional-programming-techniques-with-ruby-part-i/](http://www.sitepoint.com/functional-programming-techniques-with-ruby-part-i/), [https://docs.python.org/2/howto/functional.html](https://docs.python.org/2/howto/functional.html).

With that in mind, it's useful to think of R as supporting three ways of thinking how to organize computation:

1. Functions and lists, in the _functional_ language sense described above,  
2. Numerical computation, operations over vectors, matrices and similar mathematical structures,  
3. Operations over *data tables* (think spreadsheets in Excel, or individual tables in a database). We will start our introduction to R here.  

## Literate Programming

One last note before we get started. R has great support for [literate programming](http://en.wikipedia.org/wiki/Literate_programming), where source code that contains both code, the result of evaluating that code, and text explaining that code co-exist in a single document. This is extremely valuable in data analysis, as many choices made by data analysts are worth explaning in text, and interpretation of the results of analyses can co-exist with the computations used in that analysis. This document you are reading contains both text and code. In class, we will use [Rmarkdown](http://rmarkdown.rstudio.com/) for this purpose.

## A simple data analysis to get us going

I'm going to do a very simple analysis of Baltimore crime to show off R. We'll use data downloaded from Baltimore City's awesome open data site (this was downloaded a couple of years ago so if you download now, you will get different results). 

The repository for this particular data is here. [https://data.baltimorecity.gov/Crime/BPD-Arrests/3i3v-ibrt](https://data.baltimorecity.gov/Crime/BPD-Arrests/3i3v-ibrt) 

First, make sure you can get Rstudio running. See [this page](http://www.hcbravo.org/IntroDataSci/homeworks/rocker.html) for instructions.

We recommend that you use Rstudio for this class. Once open you will see Rstudio is organized into multiple panes. You will have a console into which you can write and execute commands directly. You will also have an editor pane where you can edit R source code. R scripts (only containing commands) are stored in `.R` files, Rmarkdown files (containing both commands and text) are stored in `.Rmd` files. You can follow along by writing commands shown here either in an `R`, or `Rmd` file and executing commands from the editor pane.

### Getting data

We've prepared the data previously into a comma-separated value file (`.csv` file). In this format, each line contains _attribute_ values (separated by commas) for one _entity_ in our dataset. Which we can download and load into our R environment.

```{r download}
if (!file.exists("BPD_Arrests.csv")) {
  download.file("http://www.hcbravo.org/IntroDataSci/misc/BPD_Arrests.csv", destfile="BPD_Arrests.csv")
}
```

The `read.csv` command is part of base R and allows you to read a dataset stored in a csv file. This function is extremely versatile, and you can read more about it by using the standard help system in R: `?read.csv`. Now, the result of running calling this function is the data itself, so, by running the function in the console, the result of the function is printed. 

### Variables and Value

To make use of this dataset we want to assign the result of calling `read.csv` (i.e., the dataset) to a variable:

```{r vars1}
arrest_tab <- read.csv("BPD_Arrests.csv", stringsAsFactors=FALSE)
```

Now we can ask what _type_ of value is stored in the `arrest_tab` variable:

```{r type}
class(arrest_tab)
```

The `data.frame` is a workhorse data structure in R. It encapsulates the idea of _entities_ (in rows) and _attribute values_ (in columns). We can ask other features of this dataset:

```{r questions}
# This is a comment in R, by the way

# How many rows (entities) does this dataset contain?
nrow(arrest_tab)

# How many columns (attributes)?
ncol(arrest_tab)

# What are the names of those columns?
colnames(arrest_tab)
```

Now, in Rstudio you can view the data frame using `View(arrest_tab)`, which can be very useful.

### Indexing

Now a basic operation in data analysis is selecting subsets of a dataset. For that we can use a few alternative options for _indexing_ into datasets.

```{r}
# to obtain the value in the first row, fifth column:
arrest_tab[1,5]

# note that indexing in R is 1-based, not 0-based, so the first row is indexed by 1

# now we want to do a bit more, so let's say we want the value in the fifth column of our dataset for the first 10 rows. For that we can use slice notation:
arrest_tab[1:10,5]

# similarly, to obtain the value in the first five columns of the first row
arrest_tab[1,1:5]

# what is the class of the value when we subset a single column?
class(arrest_tab[1:10,5])

# what is the class of the value when we subset a single row?
class(arrest_tab[1,1:5])

# what do we get with this indexing?
arrest_tab[1:10,1:5]
```

We can index any set of rows or columns by constructing _vectors_ of integers. In fact, the slice notation `:` is essentially doing that for a sequence of consecutive indices. You should think of vectors as lists of values with the same class.

If we want non-consecutive indices we have other options (e.g., the `c` function, for "concatenate")

```{r}
# non-consecutive indices using c
arrest_tab[c(2,4,7,10), 1:5]

# here's a fun one, when we call columns for a subset of rows
arrest_tab[c(2,4,7,10), ]

# there is also the `seq` function, to create sequences
arrest_tab[seq(from=1,to=10), seq(1,10)]

# that is equivalent to 
arrest_tab[1:10,1:10]

# with the `seq` function you can do more sophisticated things like select only entries in odd rows (1,3,5,7...)
head(arrest_tab[seq(from=1,to=nrow(arrest_tab),by=2), ])
```

Now, since columns have names, we can also use strings (and vectors of strings) to index data frames.

```{r}
# single column
arrest_tab[1:10, "age"]

# multiple columns
arrest_tab[1:10, c("age", "sex", "race")]
```

If we wanted a single named column from a data frame there's a special operator `$` to index:

```{r}
# first ten values of the age column
arrest_tab$age[1:10]

# EXERCISE
# try using three different ways of selecting rows 20 to 30 # of the "sex" column
```

In addition to integer indices or names, we can use vectors of logical values for indexing. 

```{r}
# rows 2,4,7 and 10 using logical indices
arrest_tab[c(FALSE,TRUE,FALSE,TRUE,FALSE,FALSE,TRUE,FALSE,FALSE,TRUE,rep(FALSE,nrow(arrest_tab)-10)),]

# now here's a fun one, if we only wanted odd rows
head(arrest_tab[c(TRUE,FALSE),])
```

The last example shows one of the most common gotchas in R. Indices are recycled. For instance if selecting rows, if you pass a logical vector that's shorter than the number of rows  in the data frame, the vector will be recycled as many times as necessary to match the number of rows in the dataset. Now, why is this useful, because a pithy index vector can let you select easily. Why is this bad, because errors in code can go easily unnoticed. So in this case, the price of ease of use is paid by the programmer by having to think a lot more carefully about their code (this is a theme in R programming...)

The utility of logical indexing is that now we can select rows based on a property of its values for a given column

```{r}
# select rows for entities younger than 21 years old
head(arrest_tab[arrest_tab$age < 21, ])

# notice that the value of expression `arrest_tab$age < 21` # is a logical vector

# select entities (arrests) occuring in Mount Washington,
# a specific neighborhood in Baltimore
head(arrest_tab[arrest_tab$neighborhood == "Mount Washington",])

# how about arrests where subjects are under 21 in Mount  Washington? 
# use a logical `and` operator
indices <- arrest_tab$age < 21 & arrest_tab$neighborhood == "Mount Washington"
```


## Exploration

R has built-in functions that help easily obtain summary information about datasets. For instance:

```{r}
summary(arrest_tab$sex)
summary(arrest_tab$race)

# well that seems problematic
# let's rename columns to correct that
colnames(arrest_tab)[3:4] <- c("race", "sex")
```

We can also ask other useful type of summaries

```{r}
# What is the average age in arrests?
mean(arrest_tab$age)

# Median age?
median(arrest_tab$age)

# what types of offenses are there
summary(arrest_tab$incidentOffense)

# what does summary looks like for continuous attributes?
summary(arrest_tab$age)
```

Combining this type of summary with our indexing strategies we learned previously we can ask more specific questions

```{r}
# What is the average age for arrests in Mount Washington?
mount_washington_index <- arrest_tab$neighborhood == "Mount Washington"

mean(arrest_tab$age[mount_washington_index])

# How about the number of arrests in Mount Washington _stratified_ by race and sex?
table(arrest_tab$race[mount_washington_index], arrest_tab$sex[mount_washington_index])

# how about a graphical summary of arrest ages in Mount Washington?
# we'll use a boxplot
boxplot(arrest_tab$age[mount_washington_index])

# can we do the same stratified by age?
boxplot(arrest_tab$age[mount_washington_index]~arrest_tab$sex[mount_washington_index])
```

This used a very useful notation in R: the tilde, `~` which we will encounter in a few different places. One way of thinking about that abstractly is, do something with this attribute, as a function (or depending on, stratified by, conditioned on) this other attribute. For instance, "plot `age` as a function of sex" in our example.

Let's write code that's a little cleaner for that last plot,
 and let's also make the plot a bit more useful by adding a title and axis labels:
 
```{r}
mount_washington_tab <- arrest_tab[mount_washington_index,]
boxplot(mount_washington_tab$age~mount_washington_tab$sex,
        main="Mt. Washington", 
        xlab="Sex", ylab="Arrest Age")
```

Here's one more useful plot:

```{r}
barplot(table(mount_washington_tab$race), 
        xlab="Number of Arrests",
        ylab="Race")
```

### Functions (and conditionals)

Now suppose we wanted to do a similar analysis for other neighborhoods. In that case we should encapsulate the summaries and plots we want to do in a function:

```{r}
analyze_neighborhood <- function(neighborhood) {
  neighborhood_index <- arrest_tab$neighborhood == neighborhood
  neighborhood_tab <- arrest_tab[neighborhood_index,]
  
  # actually, we only want to make the plots for neighborhoods where there are more arrests for whites than blacks, let's use a conditional expression
  if (sum(neighborhood_tab$race == "W") > 
      sum(neighborhood_tab$race == "B")) {
    
    boxplot(neighborhood_tab$age~neighborhood_tab$sex,
            main = neighborhood,
            xlab = "Sex", ylab="Arrest Age")
    
    barplot(table(neighborhood_tab$race),
            main = neighborhood,
            xlab = "Race", ylab="Number of Arrests")
  }
}
```

Now we can use that function to make our plots for specific neighborhoods

```{r}
analyze_neighborhood("Mount Washington")
analyze_neighborhood("Hampden")
```

## Loops

Finally, we can use a for loop to make these plots for all neighborhoods in the dataset and save them to a pdf file

```{r}
neighborhoods <- unique(arrest_tab$neighborhood)

# open a pdf file and save all plots to that file from
# this point until file is closed
pdf("neighborhood_analysis.pdf", width=6, height=6)
for (neighborhood in neighborhoods) {
  analyze_neighborhood(neighborhood)
}
# close the pdf file (plots will now be displayed in Rstudio again)
dev.off()
```

## Next up

On the intricacies of vectorized operations and loops. Matrices and other numerical computing facilities.
