---
title: "SQL Examples with Baseball"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

```{r setup_db}
library(DBI)
db <- dbConnect(RSQLite::SQLite(), dbname="data/lahman2016.sqlite")
knitr::opts_chunk$set(connection="db")
```

## Select-From-Where

Batting statistics for Washington Nationals in 2010: note the *rename*

```{sql, connection=db}
SELECT b.playerId, b.yearId, b.H, b.AB
FROM BATTING AS b
WHERE teamID = 'WAS' AND yearID = 2010
```


## Expressions

The **select** clause can contain expressions (this is paralleled by the `mutate` operation we saw previously)

```{sql, connection=db}
SELECT b.playerId, b.yearId, b.AB, 1.0 * b.H / b.AB AS BP
FROM BATTING AS b
```


## WHERE predicates

The **where** clause support a large number of different predicates and combinations thereof (this is parallel to the `filter` operation)

```{sql, connection=db}
SELECT b.playerId, b.yearID, b. teamId, b.AB, 1.0 * b.H / b.AB AS BP
FROM BATTING AS b
WHERE b.AB > 0 AND
  b.yearID > 2000 AND
  b.yearID < 2010 AND 
  b.teamID LIKE 'NY%'
```

## ORDERING

We can include ordering (parallel to `arrange`)

```{sql, connection=db}
SELECT b.playerId, b.yearID, b. teamId, b.AB, 1.0 * b.H / b.AB AS BP
FROM BATTING AS b
WHERE b.AB > 0 AND
  b.yearID > 2000 AND
  b.yearID < 2010 AND 
  b.teamID LIKE 'NY%'
ORDER BY b.AB DESC, BP DESC;
```

## Group_by and Summarize

- What it does: Partition the tuples by the group attributes (*teamId* and *yearId* in this case), and do something (*compute avg* in this case) for each group
- Number of resulting tuples == Number of groups

```{sql, connection=db}
SELECT b.teamId, b.yearId, avg(1.0 * b.H / b.AB) AS AVE_BP
FROM BATTING AS b
WHERE b.AB > 0 AND
  b.yearID > 2000 AND
  b.yearID < 2010
GROUP BY b.teamId, b.yearId
ORDER BY AVE_BP DESC
```

## Subqueries

Sometimes it's easier to nest queries like the one above into query and subquery

```{sql, connection=db}
SELECT teamID, yearID, avg(BP) AS AVG_BP
FROM (SELECT b.teamId, b.yearId, 1.0 * b.H / b.AB AS BP
      FROM BATTING AS b
      WHERE b.AB > 0 AND
        b.yearID > 2000 AND
        b.yearID < 2010)
GROUP BY teamID, yearID
ORDER BY AVG_BP DESC;
```


```{r}
dbDisconnect(db)
```